---
title: "SurvGPR Example"
author: "Aaron J. Molstad (amolstad@fredhutch.org)"
date: "5/28/2019"
output:
  html_document: default
  pdf_document: default
---
In this document, we show how to use the $\texttt{SurvGPR}$ R package. **If you encounter any errors or strange behavior, please report the issue at https://github.com/ajmolstad/SurvGPR.** First, we download the package from GitHub. 
```{r setup, include=TRUE, eval=FALSE}
library(devtools)
devtools::install_github("ajmolstad/SurvGPR")
```

Then, we create survival data from the Gaussian process regression model. 
```{r generate_data, include = TRUE, cache=TRUE}
sessionInfo()
library(SurvGPR)
library(MASS)
# set dimesions
set.seed(20)
n = 500
p = 1000
q = 2
train.inds <- sample(1:n, 300)
test.inds <- c(1:n)[-train.inds]

# generate gene expression kernel
SigmaX <- matrix(.5, nrow=p, ncol=p)
diag(SigmaX) <- 1
X <- mvrnorm(n = n, mu = rep(0, p), Sigma = SigmaX, tol = 1e-6)
Kout <- as.matrix(dist(X, method="euclidean"))
K1 <- exp(-Kout/max(Kout[train.inds,train.inds]))
K1.full <- exp(-Kout/max(Kout))

Z <- cbind(rep(1, n), matrix(rnorm(n*q), nrow=n))
beta <- c(6.2, -0.5, -1.2)
Sigma <- 2*K1.full
G <- mvrnorm(n = 1, mu = rep(0, n), Sigma = Sigma, tol = 1e-6)
	
# generate failure times 
log_time <- Z%*%beta + c(G) + rnorm(n, sd=sqrt(.5))		
C <- log(rexp(n=n, rate=1/quantile(exp(log_time), .8)))
y <- pmin(log_time, C)
time <- exp(y)
status <- 1*(y != C)
```
Note that the $M$ candidate kernels should be organized into an array of dimension $n_{\rm train} \times n_{\rm train} \times M$. Note that even if $M = 1$, this must still be entered as an array. In addition, each of the candidate the kernels should be positive definite. In the code above, the candidate kernels are stored as $\texttt{K1}$.

Next, we fit the $\texttt{SurvGPR}$ model using the following. Please note that this may take a few minutes to run. 
```{r fit_SurvGPR, include = TRUE,  cache=TRUE}
time.train <- time[train.inds]
status.train <- status[train.inds]
Z.train <- Z[train.inds,]
K.train <- array(K1[train.inds,train.inds], dim=c(length(train.inds), length(train.inds), 1))

# ------------------------
# display inputs 
# ------------------------
time.train[1:10]
status.train[1:10]
Z.train[1:10, ]
str(K.train)
```
Finally, we fit the model using the $\texttt{SurvGPR}$ function. 
```{r fit_SurvGPR2, include = TRUE , cache=TRUE}
set.seed(21)
results <- SurvGPR(time = time.train, status = status.train, Z = Z.train, K = K.train, 
                   tol = 1e-7, max.iter.MM = 100, max.iter = 100, quiet=FALSE, 
                   max.samples = 1e5, initializer = 0)
```
Checking the output, we have a list of parameter estimates and other information needed for subsequent prediction tasks. 
```{r results_print, include = TRUE , cache=TRUE}
str(results)
```

For example, we can look at the estimated variance components, which are fairly accurate in this particular dataset:  
```{r, cache=TRUE}
results$sigma2
```


With the fitted model, we can then predict for new subjects. Note that we can also get predicted survival probabilities at certain times (on the original scale, not log-scale) by specifying times in the $\texttt{times}$ argument. The first two arguments should be input as shown below: $\texttt{Z} \in \mathbb{R}^{n \times p+1}$, $\texttt{K.testt}  \in \mathbb{R}^{n \times n \times M}$ where $\texttt{train.inds}$ and $\texttt{test.inds}$ are a partition of $\left\{1, \dots, n\right\}$ where $\texttt{train.inds}$ denote the rows of $\texttt{Z}$ and row/columns of $\texttt{K}$ corresponding to the training subjects; and $\texttt{test.inds}$ corresponds to the testing subjects. 
```{r, cache=TRUE, tidy=TRUE}
K.test <- array(K1.full, dim=c(n,n, 1))
pred_results <- SurvGPR_Predict(results = results, barT = results$Tout , Z.full = Z, K.full = K.test, train.inds = train.inds, test.inds = test.inds, times = seq(0, max(time)/5, length=1000))
str(pred_results)
```
The output contains the predicted survival times on the log-scale, the testing indices, and the survival probabilities for all test set patients at the values given for \texttt{times}. We plot the true survival curve (dashed) and the estimated survival curve (solid) for two subjects below (differentiated by colors). 
```{r survPlot4, cache=TRUE, tidy=TRUE}
dim(pred_results$survFunc)
plot(x = seq(0, max(time)/5, length=1000), y = pred_results$survFunc[4,], 
     ylab="Survival probability", xlab="Survival time", type="l", ylim=c(0,1))
lines(x = seq(0, max(time)/5, length=1000), y = 1 - pnorm(log(seq(1, max(time)/5, length=1000)), mean = c(Z[test.inds[4],]%*%beta), sd = sqrt(3.5)), lty=2)
lines(x = seq(0, max(time)/5, length=1000), y = 1 - pnorm(log(seq(1, max(time)/5, length=1000)), mean = c(Z[test.inds[1],]%*%beta), sd = sqrt(3.5)), lty=2, col="red")
lines(x = seq(0, max(time)/5, length=1000), y = pred_results$survFunc[1,], lty=1, col="red")
```


We also plot our predictions versus the true log-survival times. 
```{r survPlot2, cache=TRUE, tidy=TRUE}
plot(x = pred_results$log.pred, y = log_time[pred_results$test.inds], 
     xlab="Predicted log-survival times", ylab="True log-survival times", pch=20)
abline(0,1,  lty=2)
```
